Comments on Blair and Sambanis, 2020, \"Forecasting Civil Wars: Theory and Structure in an Age of 'Big Data' and Machine Learning\", *JCR*

### 2020-06-09 paper.Rmd
maybe penultimate
version 0.9

---
title: "Comments on Blair and Sambanis, 2020, \"Forecasting Civil Wars: Theory and Structure in an Age of 'Big Data' and Machine Learning\", *JCR*"
author: 
 - Andreas Beger^[Predictive Heuristics, adbeger@gmail.com. corresponding author]
 - Richard K. Morgan^[USAID, rick.morgan2@gmail.com. The views expressed are those of the author and do not reflect the official policy or position of USAID or the US Government.]
 - Michael D. Ward^[Predictive Heuristics, Duke University, University of Washington, michael.don.ward@gmail.com]
date: "`r format(Sys.time(), '%d %B %Y')`"
thanks: Cassy L. Dorff provided helpful feedback on this project.
abstract: We examine the research protocols in Blair and Sambanis (2020).  We find that there are several important mistakes and research decisions that determine their results. Fixing these mistakes results in a reversal of their claim that theory based models of escalation are better at predicting onsets of civil war than other kinds of models.  While their model is not very theoretical, it is outperformed by several of the ad hoc, non-theoretical models they examine.  
output: 
  bookdown::pdf_document2:
    keep_tex: true
    toc: false
bibliography: 
  - "`r user <- Sys.info()[['user']]; if (user=='andybega')  '/Users/andybega/Work/whistle/master.bib' else if (user=='mdw') '/Users/mdw/Git/whistle/master.bib' else 'references.bib'`"
---

@@

library(bookdown)
library(readr)
library(dplyr)
library(tidyr)
library(tidyverse)
library(kableExtra)
library(stringr)
```
# Introduction

Blair and Sambanis (2020; hereafter B&S) argue that theory is important for creating models that have high accuracy in forecasting civil war onset.  Indeed they argue that with such theory, forecasting is more accurate than has previously been possible.  Setting aside the validity of this argument, we re-examine the empirical basis for the claims made in their article.  We find that these claims are false.  Their theory-based escalation model does not do better than the alternatives that they examine, indeed it does worse.  The reason for this reversal of their conclusion is that they have made several mistakes in their research procedure, the principal of which is that the performance results they report are based on smoothed performance curves, not the original unsmoothed curves. In addition, two of the structural alternatives to their basic escalation model were incorrectly implemented, and we also found that the scoring of their forecasts for the first half of 2016 were incorrectly performed using civil war incidence, not onset. In what follows we show the impact of these mistakes on the conclusions, we examine their understanding of prediction itself, and we explore the logic of their assertion about theory as well.

B&S claim (page 3) to show that a model informed by procedural theories of escalation and de-escalation can predict the onset of civil wars "remarkable accurately." Indeed they argue that this so-called theoretical model outperforms four other "more mechanical" alternatives. Second, they clam that the integration of structure with process are better over short forecasting windows. Third they preregistered the list of thirty countries which have the highest risk of civil war onset. They claim that such prospective predictions are rare in the literature when in fact they have been routine for many years with several prominent projects. B&S claim to be unique in assessing these forecasts. A qualitative analysis of their predictions allows them to conclude that their model is strong and a more precise test is undertaken. We will return to their analysis later, after correcting the procedural mistakes we found in their research process.

@@

@blair:sambanis:2020 (B&S hereafter) aim to examine whether theory adds more value to a forecasting model when compared to non-parametric machine learning models, and specifically random forests, whose specifications are not theory-informed. For the moment we set aside a) the logic of this hypothesis and b) whether their model has more theory than is typically found in empirical conflict models.  In short, they uphold their hypothesis that theory guided empirical research produces better conflict predictions than machine learning inspired efforts that are essentially ad hoc combinations of available variables. They arrive at this conclusion by examining the problem of predicting civil war onset. They report that a parsimonious model using a small number of covariates derived from escalation theories of conflict can forecast civil war onset better than alternative specifications based on generic covariates not specifically informed by theory including a \textit{kitchen sink} model with more than 1,000 covariates.

@@

To examine the first question, B&S compare the test set of the escalation model to four alternative models. The independent variables for the first set of analysis reported in Table 1 in the paper are all derived from the ICEWS event data, using domestic events between actors within a country. The models are:
- Escalation: a set of ten indicators, putatively drawn from a theoretical escalation model.
- Quad: ICEWS quad counts, i.e. material conflict, material cooperation, verbal conflict, verbal cooperation.
- Goldstein: -10 (conflictual) to 10 (cooperative) scores derived from the ICEWS data for interactions between the government one one side and opposition or rebel actors on the other. These are directed, thus making for four total covariates.
- CAMEO: counts for all CAMEO event codes totaling $1,159$ covariates, which are mostly zero for any country in any month. 
- Average: unweighted average of the predictions from the four models briefly described above.

@@

Our belief is that these research decisions and issues lead B&S to draw conclusion that are incorrect. The escalation model is not the best and it actually performs worse that the atheoretical, garbage can model with over $1000$ variables.  We turn to discussing these five issues below.  We hold a complete analysis that corrects all these issues until later, as there are many possible permuations of a seriatim unfolding.

@@

The most consequential issue that we found is that all AUC-ROC values reported in B&S Tables 1 and 2 are calculated using smoothed ROC curves, not the original, actual ROC curves. the data, which specifically in rare events problems like this one restricts the number of distinct true positive rate values. As we show below, the use of smoothed ROC curves to calculuate AUC has a dramatic impact on the results. To our knowledge, the norm is to calculate AUC-ROC values using original, not smoothed ROC curves. We are in fact not aware of other work that uses smoothed ROC curves for AUC calculations. It might be that there are theoretical reasons justifying the use of smoothed ROC curves over the original ROC curves; but given that this decision dramatically impacts the interpretation of the B&S results, it minimally would have warranted an explicit discussion in the paper. As it is, B&S offer no justification for using smoothed ROC curves in the AUC-ROC calculations.

@@

B&S intend that the escalation model's predictions for the test set are weighted by the PITF model predictions for the test set. B&S, however, actually weights the \textbf{test} set predictions using the PITF model predictions for the \textbf{training} set.^[See `1mo_run_escalation_weighted_PITF.R` line 4, where the PITF predictions are taken from the training data set (`train$pred_prob_plus1`). The next line is a hack extending the shorter `weight` vector with missing values to avoid a R warning when it is multiplied with the longer vector of escalation model test set predictions. Similarly in the 6-month version of this file.] This appears to be an easily corrected coding error on the part of B&S.

@@

Similarly, the "PITF Split Population" model appears to be incorrectly implemented, owing to a coding error. B&S describe it on page 20:

@@

The intention if B&S appears to have been to run two separate random forest models, one each on the low- and high-risk training data splits. The replication code does indeed run two separate random forecasts, but they are both run on the *exact same training data*, which consists of the full training data all other models are run on. The models are also identical otherwise, i.e. they use the same *x* variables and the same random forest hyper-parameter settings. The *only* difference in the models as they are implemented in the B&S replication code is due to the non-deterministic nature of the random forest model itself. If we ran both with the same random seed, they would be identical in every respect, producing identical predictions.^[Disentangling this coding error is not straightforward as it occurs over several R scripts and requires (or at least is easier to verify by) running partway through the actual replication until the objects holding the training data for the models are instantiated and can be examined. We have documented details at https://github.com/rickmorgan2/Blair-Sambanis-replication/issues/5.]

The implementation error aside, this split-population analog model is actually quite odd and does not actually replicate an analogue of the idea behind split-population modeling. Although the two RFs are trained on separate data (in our updated, fixed replication), the process of combining them actually just creates a new, larger RF using both component model's underlying decision trees. Thus, while all RF models throughout (except for one of the robustness checks) are trained with 100,000 decision trees (`ntree`), the new RF model after combination does indeed have 200,000 decision trees. Furthermore, the PITF model predictions do not impact the way the combined RF model predicts at all, not even through a binary low-/high-risk split. The split-population PITF RF model is practically speaking just another escalation model trained with N=200,000 instead of N=100,000 trees and an extra odd randomization step added to the already existing RF randomization facilities (row and column sampling for each decision tree). This does not adequately implement their research strategy.

@@

Further, the AUC-ROC values reported in the original B&S Table 2 are calculated on the basis of slightly different numbers of underlying test set cases (see Table \ref{tab:table2-N}). ROC calculations for a set of predictions can only be done on the set of cases for which both non-missing predictions and non-missing outcomes are available. Those sets differ across models (columns) for each row in B&S Table 2.

Thus a difference in AUC-ROC values for two models could be due to the fact that they were calculated on different sets of underlying cases, not because the models are systemically performing at a different level. In other words, the results for different models in B&S Table 2 are actually not comparable to one another, and any conclusions drawn from such comparison are potentially incorrect.

@@

B&S show a confusion matrix to score their 2016-H1 forecasts in Table 4. Although the forecasts are for the probability of civil war onset, in the replication code they are actually scored using the much more common incidence of civil war, i.e. including ongoing civil wars as "1"'s.

@@

By examining the code used to generated Table 4, we were able to confirm that the onset forecasts are assessed using incidence, not onset. In the file `6mo_make_confusion_matrix.do` on line 52, missing values in "incidence_civil_ns" are recoded to 1, thus reverting the onset coding of this variable back to incidence.

@@

![Replication of B&S Figure 1 with both smooth and non-smoothed ROC curves.](figures/figure1-replicated.png)

The specific predictions here also include groups of cases with identical predicted probabilities, which accounts for the unusual diagonal lines as seen in the panels on the right. In any case, with a sparse outcome like civil war onsets, the true positive rate on the *y*-axis only changes when the prediction for a observed positive case is reached. For these ROC curves, and for that matter in the basic train/test split used for 12 of the 18 rows/models in B&S Table 1, there are only 11 civil war onset cases in the test set. Thus, the ROC curves here are very step-like, with only 12 (11 positive cases plus 1 for TPR = 0) distinct *y* coordinates.  Notice also that the smoothing averages the left most almost straight line with the right most almost straight line in a monotonic way.

@@

In sum, it is not only the case that using smoothed ROC curves alters the results: the use of smoothed ROC curves to calculate AUC-ROC values clearly benefits \textbf{only} the escalation model. It does so consistently and by a considerable margin. All eight alternative models reported in B&S Tables 1 and 2 on average do not gain when using smoothed ROC curves to calculate AUC.

@@

Both the Average and CAMEO models outperform the Escalation model in almost all instances. The Goldstein model generally outperforms the Escalation model in the 6-month version. The Quad model appears to be roughly on par with the Escalation model. Thus, the original B&S conclusion that the escalation model is superior to the alternative models is completely conditional on the non-standard use of smoothed ROC curves, and overturns when using traditional AUC-ROC calculations.

@@

The key difference between Table \ref{tab:table1-nosmooth} and B&S Table 1 is whether the underlying ROC curves were original or smoothed; the test set N and coding issues did not affect this set of results.^[We should note that in our replication we consciously decided to not set RNG seeds, even though the random forest models are non-deterministic and vary slightly from run to run. As we changed the replication code to allow running in parallel, we cannot exactly reproduce the B&S results even with the same RNG seed. More importantly, the substantive interpretation of results should not depend on a specific RNG seed. We found that the escalation model's AUC-ROC values generally fluctuate no more than 0.01 (see https://github.com/rickmorgan2/Blair-Sambanis-replication/blob/master/rep_nosmooth/variance.md), and thus are confident that the patterns we see are robust to the initial RNG state.] B&S's original results and interpretation regarding the superiority of the B&S model over alternative ICEWS models, including the 1,000+ covariate CAMEO model, are thus entirely conditional on the use of smoothed AUC-ROC. Even the Quad model is generally as good as the escalation model in all implementations, and is frequently quite a bit better.

@@

Generally, these comparisions require the same cases.  Therefore, we use predictions for common joint subset of nonmissing predictions The resulting numbers of cases for each model are shown in the bottom portion of the table. Since the sets of incomplete cases in the original version above do not completely overlap themselves, the common subset for all models is actually slightly smaller than the smallest N in the top portion of the table.

@@

Table \ref{tab:table2-fixed} shows our replication of B&S Table 2 with (1) regular, not smoothed, AUC-ROC, (2) fixed "Weighted by PITF" and "PITF Split-Population" models, and (3) AUC-ROC values computed on the common, joint subset of tests cases that all models have non-missing predictions for. Table \ref{tab:table2-full} in the appendix shows AUC-ROC values for both smoothed and non-smoothed versions, and both the original, model-varying test cases sets and our common joint subset.

@@

The most straightforward method of incorporating the annual, structural PITF variables--adding them to the predictors of the Escalation RF model--strictly outperforms the Escalation Only model. Note that the two other combination models considered are both non-standard and that the "PITF Split Population" model does not in fact actually incorporate structural information at all. We thus conclude that adding structural variables improves predictive performance.

@@

B&S report confusion matrices for their forecasts for the first half of 2016 (2016-H1) in their Table 4. To create the confusion matrices, B&S treat the 30 highest ranked predictions as positive predictions ("1s") and the rest as negative predictions ("0s"). We replicate their Table 4 in the top portion of \ref{tab:table4}. There are two confusion matrices for slightly different codings of outcomes in 2016-H1, under the corresponding "Assuming Persistence" and "Assuming Change" headings. We should note that the "Assuming Persistence" corresponds to the values in the replication data; the small variations for the "Assuming Change" version are hand-coded in the replication code file that generates the confusion matrices and appear to be subjective assessments of B&S.

We can see that the original table presents 15 or 16 positive cases for 2016-H1, depending on the dependent variable coding variation. This corresponds to a positive rate of around 9.5% for the first half of 2016 data. In contrast, the corresponding 6-month version of the data from 2001 to 2015, with 30 half-years, has in total 20 civil war onset events, for a much lower positive rate of around 0.5%. Since the positive event rate in the confusion matrices far exceeds the rate of observed civil war onsets in both the training and test data, this suggests that the forecasts were erroneously assessed using civil war incidence, not onset. As we showed above, by examining the replication code we were able to verify that the forecasts were scored using civil war incidence, i.e. including ongoing civil wars, rather than civil war onset years only.

@@

The correct confusion matrices when using observed onset (or the lack of it) are shown in the second part of Table \ref{tab:table4}. In the default "Assuming Persistence" coding, there are no civil war onsets in the data for 2016-H1. Thus, the recall values is undefined, while the precision is 0/30 = 0, compared to reported recall and precision values of 13/15 = 0.87 and 13/30 = 0.43. The alternative coding ("Assuming Change") produces 2 civil war onsets. Recall is 1 compared to 14/16 = 0.88 before, and precision is 2/30 = 0.07 instead of 14/30 = 0.47.

Another, minor issue or rather coding error, is related to using a lead version of the DV. With the lead version of the DV, "incidence_civil_ns_plus1", which is what the models are predicting, the predicted value for 2016-H1 actually indicates the risk of civil war onset in 2016-H2. In the Table 4 script referenced above, the 2016-H1 predictions (for 2016-H2) are assessed using the raw DV, "incidence_civil_ns", not the lead version. Essentially, the forecasts for 2016-H2 are assessed using observed outcomes for 2016-H1. In this case it doesn't make a difference since both the raw DV and lead version for 2016-H1 do not have any positive values.

@@

The B&S approach that is advocated is to use theory to guide prediction. But theory as used by B&S is an ambiguous and undefined concept.  It is not a procedure.  What they actually do is to create a model with ten right hand side variables that is supposed to capture a complicated repression-dissent dynamic. There is a wide-ranging literature on this dynamic that they do not rely on to construct their model.  As such their baseline comparison is a poor standard bearer for strong theory. Moreover, they misunderstand the use of ICEWS event data in current research. They claim that most uses to date have focused on the quad categories, but this ignores a wide swath of literature (Steinert-Threlkeld APSR 2017) and Metternich et al (AJPS 2013) that uses a specific action---such as protest---defined in the CAMEO ontology. In the AJPS article on Thailand, Metternich et al hand coded, for example, every actor in Thailand and focused on an analysis of how those interact.

We encountered several issues in the code underlying the B&S analysis. The issues we encountered are not subjective modeling choices. When we fix these issues and perform an updated analysis, the B&S conclusions are all essentially overturn. In other words, B&S findings are based on a faulty analysis, and invalid.

Using the same analysis B&S intend to use, we in fact find that:
- the theory-driven escalation model is outperformed both by the low-effort 1,160 predictor all CAMEO model and the Average ensemble model
- structural variables substantially improve the escalation model's performance when added to the pool of predictors on which the underlying random forest model is based.

Blair & Sambanis have focused attention on comparing forecasting models and forecasts in the civil war domain.  As social science becomes more adept at predictive analysis this will doubtless be of increasing importance.  However, these comparisons must be done carefully to ensure that correct inferences are drawn.  We continue to think that theory is overrated (@ward:2016), and that machine learning and big data will allow us to learn new things, but it will be interesting to see how the evidence for these claims are adjudicated with additional usage and careful evaluation.

>>>>

---
title: "Comments on Blair and Sambanis, 2020, \"Forecasting Civil Wars: Theory and Structure in an Age of 'Big Data' and Machine Learning\", *JCR*"
author: 
 - Andreas Beger^[Predictive Heuristics, adbeger@gmail.com. corresponding author]
 - Richard K. Morgan^[VDEM, rick.morgan2@gmail.com.]
 - Michael D. Ward^[Predictive Heuristics, Duke University, University of Washington, michael.don.ward@gmail.com]
date: "`r format(Sys.time(), '%d %B %Y')`"
thanks: Cassy L. Dorff and Shahryar Minhas both provided helpful feedback on this project. All the code and several additional analyses can be found at our replication archive.
abstract: We examine the research protocols in Blair and Sambanis (2020).  We find that there are several important mistakes and research decisions that determine their results. Fixing these mistakes results in a reversal of their claim that theory based models of escalation are better at predicting onsets of civil war than other kinds of models.  While their model is not very theoretical, it is outperformed by several of the ad hoc, putatively non-theoretical models they devise and examine.  
output: 
  bookdown::pdf_document2:
    keep_tex: true
    toc: false
# spacing: double
bibliography: 
  - "`r user <- Sys.info()[['user']]; if (user=='andybega')  '/Users/andybega/Work/whistle/master.bib' else if (user=='mdw') '/Users/mdw/Git/whistle/master.bib' else 'references.bib'`"
---

@@

library(bookdown)
library(readr)
library(dplyr)
library(tidyr)
library(tidyverse)
library(kableExtra)
library(stringr)
```
# Introduction
 @blair:sambanis:2020 (hereafter B\&S) argue that theory is essential for creating models that have high accuracy in forecasting civil war onset. Indeed they assert that with such theory, forecasting is more accurate than has previously been possible. Setting aside the validity of this argument, we re-examine the empirical basis for the claims made in their article. We find that these claims are false. Their theory-based escalation model does not do better than the alternatives that they examine. Indeed, it does worse. The reason for this reversal of their conclusion is that they have made several mistakes in their research procedure. The performance results they report are based on smoothed performance curves, not the original unsmoothed curves. Two of the structural alternatives to their basic escalation model were incorrectly implemented. We also found that the scoring of their forecasts for the first half of 2016 were incorrectly performed using civil war incidence, not onset. In what follows, we show the impact of these mistakes on the conclusions.

B\&S claim (page 3) to show that a model informed by procedural theories of escalation and de-escalation can predict the onset of civil wars ``remarkable accurately.'' Indeed they argue that this so-called theoretical model outperforms four other ``more mechanical'' alternatives. Second, they claim that the integration of structure with process is better over short forecasting windows. Third, they preregistered the list of thirty countries that have the highest risk of civil war onset. They claim that such prospective predictions are rare in the literature when, in fact, they have been routine for many years with several prominent projects. B\&S claim to be unique in assessing these forecasts. A qualitative analysis of their predictions allows them to conclude that their model is robust, and a more precise test is undertaken. We will return to their analysis later, after correcting the procedural mistakes we found in their research process.

@@

Blair and Sambanis (2020) (B\&S hereafter) aim to examine whether 
theory adds more value to a forecasting model when compared to non-parametric machine learning models, and specifically random forests, whose specifications are not theory-informed. For the moment, we set aside a) the logic of this hypothesis and b) whether their model has more theory than is typically found in empirical conflict models. In short, they uphold their assumption that theory-guided empirical research produces better conflict predictions than machine learning inspired efforts that are necessarily ad hoc combinations of available variables. They arrive at this conclusion by examining the problem of predicting civil war onset. They report that a parsimonious model using a small number of covariates derived from escalation theories of conflict can forecast civil war onset better than alternative specifications based on generic covariates not specifically informed by theory including a \textit{kitchen sink} model with more than 1,000 covariates.

@@

To examine the first question, B&S compare the test set of the escalation model to four alternative models. The independent variables for the first set of analysis reported in Table 1 in the paper are all derived from the ICEWS event data, using domestic events between actors within a country. The models are:
- Escalation: a set of ten indicators, putatively drawn from a theoretical escalation model.
- Quad: ICEWS quad counts, i.e. material conflict, material cooperation, verbal conflict, verbal cooperation.
- Goldstein: -10 (conflictual) to 10 (cooperative) scores derived from the ICEWS data for interactions between the government on one side and opposition or rebel actors on the other. These are directed, thus making for four total covariates.
- CAMEO: counts for all CAMEO event codes totaling $1,159$ covariates, which are mostly zero for any country in any month. 
- Average: unweighted average of the predictions from the four models briefly described above.

@@

We believe that these research decisions and issues lead B\&S to incorrect conclusions. The escalation model is not the best, and it actually performs worse than the atheoretical, garbage can model with over \(1000\) variables. We discuss these five issues below. We defer a complete analysis that corrects all these issues until later, as there are many possible permutations of a seriatim unfolding.

@@

The most consequential issue that we found is that all AUC-ROC values reported in B&S Tables 1 and 2 are calculated using smoothed ROC curves, not the original, actual ROC curves. The data in rare events problems like this one restricts the number of distinct true positive rate values. As we show below, the use of smoothed ROC curves to calculate AUC has a dramatic impact on the results. To our knowledge, the norm is to calculate AUC-ROC values using original, not smoothed ROC curves. In fact, we are not aware of other work that uses smoothed ROC curves for AUC calculations. It might be that theoretical reasons are justifying the use of smoothed ROC curves over the original ROC curves; but given that this decision dramatically impacts the interpretation of the B&S results, it minimally would have warranted an explicit discussion in the paper. As it is, B&S offer no justification for using smoothed ROC curves in the AUC-ROC calculations.

Smoothing of ROC curves is well known.  Basically, the smoothed curves transform the false positive rate and the false negative rates jointly in a monotone fashion.  The assumption is that there is a monotone transformation that produces normal distributions for the sensitivity and specificity. It is well-known that there is substantial bias that can be introduced by bias. Indeed, bootstrapping is often used as a diagnostic to see the amount of bias introduced by smoothing.

@@

B&S intend that the escalation model's predictions for the test set are weighted by the PITF model predictions for the test set. B&S, however, actually weights the \textbf{test} set predictions using the PITF model predictions for the \textbf{training} set.^[See `1mo_run_escalation_weighted_PITF.R` line 4, where the PITF predictions are taken from the training data set (`train$pred_prob_plus1`). The next line is a hack extending the shorter `weight` vector with missing values to avoid a R warning when it is multiplied with the longer vector of escalation model test set predictions. Similarly, in the 6-month version of this file.] This appears to be an easily corrected coding error on the part of B&S.

@@

The "PITF Split Population" model appears to be incorrectly implemented, owing to a coding error. B&S describe it on page 20:

@@

The intention of B&S appears to have been to run two separate random forest models, one each on the low- and high-risk training data splits. The replication code does indeed run two separate random forecasts, but they both utilize the *exact same training data*, which consists of the full training data from all other models. The models are also identical otherwise, i.e. they use the same *x* variables and the same random forest hyper-parameter settings. The *only* difference in the models as they are implemented in the B&S replication code is due to the non-deterministic nature of the random forest model itself. If we ran both with the same random seed, they would be identical in every respect, producing identical predictions.^[Disentangling this coding error is not straightforward as it occurs over several R scripts and requires (or at least is easier to verify by) running partway through the actual replication until the objects holding the training data for the models are instantiated and can be examined. We have documented details at https://github.com/rickmorgan2/Blair-Sambanis-replication/issues/5.]

The implementation error aside, this split-population analog model is quite odd and does not replicate the idea behind split-population modeling. Although the two RFs are trained on separate data (in our updated, fixed replication), the process of combining them actually just creates a new, larger RF using both component model's underlying decision trees. Thus, while all RF models throughout (except for one of the robustness checks) are trained with 100,000 decision trees (`ntree`), the new RF model after combination does have 200,000 decision trees. Furthermore, the PITF model predictions do not impact how the combined RF model predicts at all, not even through a binary low-/high-risk split. The split-population PITF RF model is practically speaking just another escalation model trained with N=200,000 instead of N=100,000 trees and an extra odd randomization step added to the already existing RF randomization facilities (row and column sampling for each decision tree). This does not adequately implement their research strategy.

@@

Further, the AUC-ROC values reported in the original B&S Table 2 are calculated based on slightly different numbers of underlying test set cases (see Table \ref{tab:table2-N}). ROC calculations for a set of predictions can only be done on the set of cases for which both non-missing predictions and non-missing outcomes are available. Those sets differ across models (columns) for each row in B&S Table 2.

Thus a difference in AUC-ROC values for two models could be because they were calculated on different sets of underlying cases, not because the models are systemically performing at a different level. In other words, the results for different models in B&S Table 2 are not comparable to one another, and any conclusions drawn from such comparison are potentially incorrect.

@@

B&S present a confusion matrix to score their 2016-H1 forecasts in Table 4. Although the forecasts are for the probability of civil war onset, in the replication code, they are actually scored using the much more commonly used incidence of civil war, i.e. including ongoing civil wars as "1"'s.

@@

By examining the code used to generated Table 4, we were able to confirm that the onset forecasts are assessed using incidence, not onset. In the file `6mo_make_confusion_matrix.do` on line 52, missing values in "incidence_civil_ns" are recoded to 1, thus reverting the onset coding of this variable to incidence.

@@

![Replication of B&S Figure 1 with both smooth and non-smoothed ROC curves.](figures/figure1-replicated.png)
The specific predictions also include groups of cases with identical predicted probabilities, which accounts for the unusual diagonal lines seen in the panels on the right. In any case, with a sparse outcome like civil war onsets, the true positive rate on the *y*-axis only changes when the prediction for an observed positive case is reached. For these ROC curves, and for that matter in the fundamental train/test split used for 12 of the 18 rows/models in B&S Table 1, there are only 11 civil war onset cases in the test set. Thus, the ROC curves here are very step-like, with only 12 (11 positive cases plus 1 for TPR = 0) distinct *y* coordinates.  Notice also that the smoothing averages the left-most almost straight line with the right-most almost straight line in a monotonic way.

@@

In sum, it is not only the case that using smoothed ROC curves alters the results: the use of smoothed ROC curves to calculate AUC-ROC values benefits \textbf{only} the escalation model. It does so consistently and by a considerable margin. All eight alternative models reported in B&S Tables 1 and 2, on average, do not gain when using smoothed ROC curves to calculate AUC.

@@

Both the Average and CAMEO models outperform the Escalation model in almost all instances. The Goldstein model generally outperforms the Escalation model in the 6-month version. The Quad model appears to be roughly on par with the Escalation model. Thus, the original B&S conclusion that the escalation model is superior to the alternative models is entirely conditional on the non-standard use of smoothed ROC curves and overturns when using traditional AUC-ROC calculations.

@@

The key difference between Table \ref{tab:table1-nosmooth} and B&S Table 1 is whether the underlying ROC curves were original or smoothed; the test set N and coding issues did not affect this set of results.^[We should note that in our replication we consciously decided to not set RNG seeds, even though the random forest models are non-deterministic and vary slightly from run to run. As we changed the replication code to allow running in parallel, we cannot exactly reproduce the B&S results even with the same RNG seed. More importantly, the substantive interpretation of results should not depend on a specific RNG seed. We found that the escalation model's AUC-ROC values generally fluctuate no more than 0.01 (see https://github.com/rickmorgan2/Blair-Sambanis-replication/blob/master/rep_nosmooth/variance.md), and thus are confident that the patterns we see are robust to the initial RNG state.] B&S's original results and interpretation regarding the superiority of the B&S model over alternative ICEWS models, including the 1,000+ covariate CAMEO model, are thus entirely conditional on the use of smoothed AUC-ROC. Even the Quad model is generally as good as the escalation model in all implementations and is frequently quite a bit better.

@@

Generally, these comparisons require the same cases.  Therefore, we use predictions for a common joint subset of non-missing predictions. The resulting numbers of cases for each model are shown in the bottom portion of the table. Since the sets of incomplete cases in the original version above do not entirely overlap themselves, the common subset for all models is slightly smaller than the smallest N in the top portion of the table.

@@

Table \ref{tab:table2-fixed} shows our replication of B&S Table 2 with (1) regular, not smoothed, AUC-ROC, (2) fixed "Weighted by PITF" and "PITF Split-Population" models, and (3) AUC-ROC values computed on the common, joint subset of tests cases for which all models have non-missing predictions. Table \ref{tab:table2-full} in the appendix shows AUC-ROC values for both smoothed and non-smoothed versions, and both the original, model-varying test cases sets and our common joint subset.

@@

The most straightforward method of incorporating the annual structural PITF variables--adding them to the predictors of the Escalation RF model--strictly outperforms the Escalation Only model. Note that the two other combination models considered are both non-standard and that the "PITF Split Population" model does not incorporate structural information at all. We thus conclude that adding structural variables improves predictive performance.

@@

B&S report confusion matrices for their forecasts for the first half of 2016 (2016-H1) in Table 4. To create the confusion matrices, B&S treat the 30 highest ranked predictions as positive predictions ("1s") and the rest as negative predictions ("0s"). We replicate their Table 4 in the top portion of \ref{tab:table4}. There are two confusion matrices for slightly different codings of outcomes in 2016-H1, under the corresponding "Assuming Persistence" and "Assuming Change" headings. We should note that the "Assuming Persistence" corresponds to the values in the replication data; the small variations for the "Assuming Change" version are hand-coded in the replication code file that generates the confusion matrices and appear to be subjective assessments of B&S.

We can see that the original table presents 15 or 16 positive cases for 2016-H1, depending on the dependent variable coding variation. This corresponds to a positive rate of around 9.5% for the first half of 2016 data. In contrast, the corresponding 6-month version of the data from 2001 to 2015, with 30 half-years, has in total 20 civil war onset events, for a much lower positive rate of around 0.5%. The positive event rate in the confusion matrices far exceeds the rate of observed civil war onsets in both the training and test data. This suggests that the forecasts were erroneously assessed using civil war incidence, not onset. By examining the replication code we were able to verify that the forecasts were scored using civil war incidence, i.e., including ongoing civil wars, rather than civil war onset years only.

@@

The correct confusion matrices when using observed onset (or the lack of it) are shown in the second part of Table \ref{tab:table4}. In the default "Assuming Persistence" coding, there are no civil war onsets in the data for 2016-H1. Thus, the recall values is undefined, while the precision is 0/30 = 0, compared to reported recall and precision values of 13/15 = 0.87 and 13/30 = 0.43. The alternative coding ("Assuming Change") produces two civil war onsets. Recall is 1.0 compared to 14/16 = 0.88 before, and precision is 2/30 = 0.07 instead of 14/30 = 0.47.

Another, minor issue or rather coding error, is related to using a lead version of the DV. With the lead version of the DV, "incidence_civil_ns_plus1", which is what the models are predicting, the predicted value for 2016-H1 indicates the risk of civil war onset in 2016-H2. In the Table 4 script referenced above, the 2016-H1 predictions (for 2016-H2) are assessed using the raw DV, "incidence_civil_ns", not the lead version. Essentially, the forecasts for 2016-H2 are assessed using observed outcomes for 2016-H1. In this case, it doesn't make a difference since both the raw DV and lead version for 2016-H1 do not have any positive values.

@@

The B&S advocate the use theory to guide prediction. But theory as used by B&S is an ambiguous and undefined concept.  It is not a procedure.  They actually create a model with ten right hand side variables that are supposed to capture a complicated repression-dissent dynamic. There is wide-ranging literature on this dynamic that could justify many specifications.  As such, their baseline comparison is an unfortunate standard bearer for strong theory.

Moreover, they misunderstand the use of ICEWS event data in current research. They claim that most applications to date have focused on the quad categories, but this ignores a wide swath of literature (@steinert-threlkeld:2017 and @metternich:etal:2013) that uses a specific action---such as protest---defined in the CAMEO ontology. In their article on conflict in Thailand, Metternich et alia hand-coded, for example, every actor in Thailand and focused on an analysis of how those interact.

We encountered several issues in the code underlying the B&S analysis. The problems we encountered are not subjective modeling choices. When we fix these issues and perform an updated analysis, the B&S conclusions are all essentially overturned. In other words, B&S findings are based on a faulty analysis, and invalid.

Using the same analysis B&S intend to use, we, in fact, find that:
- the theory-driven escalation model is outperformed both by the low-effort 1,160 predictor all-CAMEO model. The Average ensemble model
- structural variables substantially improve the escalation model's performance when added to the pool of predictors on which the underlying random forest model is based.

Blair & Sambanis have focused attention on comparing forecasting models and forecasts in the civil war domain.  As social science becomes more adept at predictive analysis, this will doubtless be of increasing importance.  However, these comparisons must be made carefully to ensure those correct inferences are drawn.  We continue to think that theory is overrated (@ward:2016), and that machine learning and big data will allow us to learn new things, but it will be interesting to see how the evidence for these claims are adjudicated with additional usage and careful evaluation.

There is a fairly long tradition of making predictions in academic and policy work on conflict. Most of these are not mentioned by B&S.  We would include work from the last century including @choucri:1974, @andriole:young:1977, @choucri:robinson:1979, @freeman:job:1979,@hopple:andriole:1984, @schrodt:1991, and others. 
The current century has even more work that can't be overlooked or combined into a theory barren classification
@obrien:2002,
@buhaug:gleditsch:2008, 
@bennett:stam:2009,
@rost:etal:2009, 
@obrien:2010, 
@weidmann:ward:2010, 
@weidmann:ward:2010, 
@ward:etal:2010,
@goldstone:etal:2010, 
@bdm:2011, 
@brandt:etal:2011, 
@hegre:etal:2011, 
@Brandt_etal_2011, 
@oloughlin:etal:2012, 
@hegre:etal:2013, @shellman:etal:2013, 
@ward:etal:2013, 
@brandt:etal:2014, 
@buhaug:etal:2014, 
@ramakrishnan:etal:2014, 
@pilster:bohmelt:2014, 
@chadefaux:2017, 
@chiba:etal:2015,
@mellers:etal:2015, 
@tetlock:gardner:2015, 
@hegre:etal:2016, 
@schleussner:etal2016, @ward:2016, 
@beger:ward:2017, 
@chiba:gleditsch:2017,
@colaresi:mahmood:2017, 
@hegre:etal:2017, 
@tetlock:etal:2017, 
@witmer:etal:2017, 
@kupilik:witmer:2018, 
@hegre:etal:2019, 
@hegre:etal:2019.
Indeed this scholarship has brought us to the point where predictive analysis of conflict studies is not unknown in the scholarly and policy arenas.  Most, if not all of these authors, would probably claim some theoretical motivation. The a-theoretical case is a counterfactual that either includes everything, or nothing, as aruged in @ward:2016. In the B&S article it serves as a strawman. Unfortunately for B&S that strawman outperforms their putatively theoretical model.^[Important statistical developments have also added to this research vector, including especially 
@gneiting:raftery:2005, 
@montgomery:etal:PS:2012, and
@mhw:2012.]
